---
title: "R to Python"
author: "Leonardo Uchoa"
date: "3/31/2020"
output: 
  pdf_document:
    toc: true
    number_sections: true
header-includes:
- \usepackage{amsbsy}
- \usepackage{amsmath}
- \usepackage{float}
- \usepackage{graphicx}
---

\newpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# The table

That's my R to python port. It's intended to make my approach to learning python faster and its mostly composed of data wrangling routine tools. Many of those are already listed in other sources.

\begin{table}[H]
\centering
\begin{tabular}{lr}
R & Python \\
\hline
dim & df.shape (pd) \\
str & df.dtypes / df.info (pd) \\
unique & np.unique (np) \\
summary & df.describe (pd) \\
\texttt{group\_by} & df.groupby (pd) \\
count & \verb|df.value_count| (pd) \\
----- & np.bincount (np) \\
apply & df.apply (pd) \\
if.else & df.where[case,true,false] (pd)  \\
table & pd.crosstab \\
mutate(df, c=a-b) & df.assign(c=df['a']-df['b']) (pd) \\
colSums(is.na()) & df.isnull().sum() (pd) \\
na.omit & df.dropna(axis=X) (pd) \\
*imputation* & df.fillna(df.mean()) (pd) \\
colnames() <- & df.colnames (pd) \\
\end{tabular}
\end{table}

Abreviations

\begin{table}[H]
\centering
\begin{tabular}{lr}
Module Abreviation & Module \\
\hline
pd & Pandas \\
np & Numpy
\end{tabular}
\end{table}

\newpage 

# Usefull examples

These are ilustrations of the commands I use the most when analysing data in R and some additional because they're different from what I'm used to do in R.

```{r}
library(reticulate)
use_python("/home/leonardo/anaconda3/bin/python")
```

```{python}
import numpy as np
import pandas as pd
```

## Counting per column

```{python}

df = pd.DataFrame(np.random.randint(0, 2, (10, 4)), columns=list('abcd'))
df.apply(pd.Series.value_counts)
```

## Categorical data encoding

```{python}

df = pd.DataFrame([
['green', 'M', 10.1, 'class2'],
['red', 'L', 13.5, 'class1'],
['blue', 'XL', 15.3, 'class2']])
df.columns = ['color', 'size', 'price', 'classlabel']

df
```

In both approaches bellow we use a dictionary the create the mapping identifier for the `map` method. Remember that according to [w3schools](https://www.w3schools.com/python/python_dictionaries.asp) a dictionary is 

>*A dictionary is a collection which is unordered, changeable and indexed. In Python dictionaries are written with curly brackets, and they have keys and values.*

### Encoding ordinals - create labels manually

```{python}
#create the dict mapping from ordinal to integer
size_mapping = {'XL': 3,'L': 2,'M': 1}

#use map to in the desired column get the mapped values
df['size'] = df['size'].map(size_mapping)
```

### Encoding nominals - creating labels automatically

```{python}
class_mapping = {label: idx for idx, label in enumerate(np.unique(df['classlabel']))}
```

Now what that command is doing is looping through the iterators `idx` and `label` (created by the `enumerate` function) in the unique values of the `classlabel` column and assigning both to `label` and `idx`. Let's see

```{python}
print(list(
    enumerate(np.unique(df['classlabel']))
  ))
```

So iterating through the list we get to assign "class1"/"class2" to `label` and 0/1 to `idx`\footnote{Note the inversion in `label: idx for idx, label` }. Finally the last step to map

```{python}
df['classlabel'] = df['classlabel'].map(class_mapping)
df
```

Want to get the mapping backwards? Access the `items` method in the `class_mapping` object and loop again

```{python}
inv_class_mapping = {a:b for b,a in class_mapping.items()}
df['classlabel'] = df['classlabel'].map(inv_class_mapping)
df
```


**Ps.: There's also an object in `skitlearn` module `preprocessing` that does this: `LabelEncoder`**

# References

[1]. Pandas: https://pandas.pydata.org/pandas-docs/stable/getting_started/comparison/comparison_with_r.html#quick-reference

[2]. Raschka, S. and Mirjalili, V., 2019. Python Machine Learning. Birmingham: Packt Publishing, Limited.